---
title: "Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lsgm)
set.seed(37)
```


## Simulation

In this vignette we simulate length selected von Bertalanffy length at age data with aging errors,
and compare the effect of fitting with `gnls`, `lsVB` ignoring ageing error, and `lsVB` allowing for
ageing error.

### Fishable Population

The function `rfage` simulates the age of `n` fished individuals assuming perfect selectivity,
constant natural mortality `M`, constant fishing mortality `F` within a fishing season that extends
over the fraction `a` to `b` of each year, where `rs` is the vector of numbers of recruits in
previous years.
```{r}
rfage <- function(n,M,F,a,b,rs) {
  ## Order recruits into the past
  rs <- rev(rs)
  amax <- length(rs)
  ## Cumulative mortality
  xs <- c(0,c(a,b)+rep(0:(amax-1),each=2),amax)
  ys <- F*c(0,c(0,1)+rep(0:(amax-1),each=2),amax)+M*xs
  ## Distribution function
  zs <- 1-exp(-ys)
  ## Numbers caught each season
  ws <- diff(zs)[seq(from=2,by=2,length.out=amax)]
  ns <- as.vector(rmultinom(1,n,ws*rs))
  ## Uniform deviates within each season
  us <- unlist(mapply(runif,ns,zs[seq(from=2,by=2,length=amax)],zs[seq(from=3,by=2,length=amax)]))
  ## Inverse probability transform
  approx(ys,xs,-log(1-us))$y
}
```


Simulate ages of the population of fish available to be caught within the fishing season, assuming a
log Normal distribution of numbers of recruits.
```{r}
N <- 1000000
M <- 0.05
F <- 0.07
season <- c(0.25,0.75)
maxAge <- 36
recruits <- rlnorm(maxAge,12,1)
df.pop <- data.frame(age=rfage(N,M,F,season[1],season[2],recruits))
hist(df.pop$age,8*maxAge)
```

Simulate the lengths of the fishable population, assuming a von Bertalanffy growth model in which
observed lengths are Normally distributed about the expected length with constant coefficient of
variation.
```{r}
Linf <- 1700
K <- 0.08
a0 <- -0.8
cv <- 0.12
df.pop <- within(df.pop,{
  mu <- Linf*(1-exp(-K*(age-a0)))
  length <- rnorm(length(mu),mu,cv*mu)
})
```

### Catch

The catch is a sample of the fishable population in which the selection probability is determined by
the gear selectivity.

Define a double Normal plateau selectivity function
```{r}
mkSelectivity <- function(xa,xb,sa,sb,ya,yb,xmin,xmax) {
  join <- function(x0,x) 1/(1+exp(-20*(x-x0)/(1+abs(x-x0))))
  normal <- function(mu,s,x) exp(-(x-mu)^2/(2*s^2))
  limb <- function(xp,s,xr,yr,x)  yr+(1-yr)*(normal(xp,s,x)-normal(xp,s,xr))/(1-normal(xp,s,xr))

  ## The selectivity
  function(x) {
    ja <- join(xa,x)
    jb <- join(xb,x)
    limb(xa,sa,xmin,ya,x)*(1-ja)+ja*((1-jb)+jb*limb(xb,sb,xmax,yb,x))
  }
}
selectivity <- mkSelectivity(500,800,120,350,0.01,0.15,0,2000)
plot(0:2000,selectivity(0:2000),type="l",xlab="Length",ylab="Selectivity")
```

Subsample the fishable population to generate the landed catch, with selection probabilities
determined by the gear selectivity
```{r}
df.catch <- df.pop[sample(nrow(df.pop),10000,prob=selectivity(df.pop$length)),]
plot(length~age,data=df.catch,pch=16,col=grey(0.5,0.1))
```


### Length Bin Sampling

The landed catch is subsampled, stratifying by length bin, to generate the sample to be aged.
Divide the catch into length bins and sample at most `n` individuals from each bin, treating the top
bin as 2000mm or greater.
```{r}
n <- 5
breaks <- seq(200,2000,10)
df.sample <- local({
  ks <- unlist(tapply(1:nrow(df.catch),
                      cut(df.catch$length,c(0,breaks,Inf)),
                      function(ks) sample(ks,pmin(n,length(ks)))))
  df.catch[ks,]
})
```

Compute the sampling fractions for each bin
```{r}
frac <- (tabulate(.bincode(df.sample$length,c(0,breaks,Inf)),nbin=length(breaks)+1)/
           pmax(1,tabulate(.bincode(df.catch$length,c(0,breaks,Inf)),nbin=length(breaks)+1)))
```


### Age Error

Create a matrix of conditional probabilities $p_{kK} = p(K \,|\, k)$ of observing age $K$ given actual age $k$
```{r}
ages <- 0:maxAge
prAge <- matrix(0,length(ages),length(ages),dimnames=list(ages,ages))
for(k in 1:nrow(prAge))
  prAge[k,] <- dbeta((ages+1)/(maxAge+2),100*k/(maxAge+2),100*(1-k/(maxAge+2)))
prAge <- prAge/rowSums(prAge)
prAge <- diag(1,length(ages),length(ages))
```

Use this matrix to generate the observed ages (otolith counts) in the sample
```{r}
df.sample$age0 <- df.sample$age
for(k in seq_along(ages)) {
  ks <- ages[k]==floor(df.sample$age0)
  df.sample$age[ks] <- sample(ages,sum(ks),replace=TRUE,prob=prAge[k,])
} 
df.sample$ageOffset <- mean(season)
plot(length~age,data=df.sample,pch=16,col=grey(0.5,0.1))
```

Simulate a "reference collection" of otoliths and use this to calculate the ageing error matrix with entries $p_{kK} = p(k \,|\, K)$
```{r}
ref <- data.frame(actual=rep(ages,each=100))
ref$observed <- double(nrow(ref))
for(k in seq_along(ages)) {
  ks <- ages[k]==ref$actual
  ref$observed[ks] <- sample(ages,sum(ks),replace=TRUE,prob=prAge[k,])
}
ageErr <- table(factor(ref$actual,ages),factor(ref$observed,ages))
ageErr <- ageErr %*% diag(1/colSums(ageErr))
```


## Growth MOdelling

### NLME 

Fit a von Bertalanffy model ignoring length selectivity and ageing error
```{r}
library(nlme)
fit.nlme <- gnls(length ~ Linf*(1-exp(-K*(age+ageOffset-t0))),
                 weight=varPower(fixed=1),
                 start=list(Linf=1600,K=0.05,t0=-1.8),
                 control=gnlsControl(nlsTol=0.01,returnObject=TRUE),
                 data=df.sample)
summary(fit.nlme)
```

The confidence intervals for the model parameters are
```{r}
confint(fit.nlme)
```

Plot the fitted growth curve
```{r}
d.nlme <- data.frame(age=seq(0,maxAge,0.1),ageOffset=mean(season))
d.nlme$length <- predict(fit.nlme,d.nlme)
plot(length~I(age+ageOffset),data=df.sample,pch=16,col=grey(0.5,0.1),ylim=c(0,1.2*Linf),xlab="age")
lines(length~age,data=d.nlme,col="firebrick")
```


### LGSM - no ageing error

Fit a von Bertalanffy model allowing for length selection, but ignoring ageing area
```{r}
library(lsgm)
ageErr0 <- diag(1,nrow(ageErr),ncol(ageErr))
fit.lsgm <- lsVB(df.sample$age,df.sample$length,ageErr0,breaks,frac,
                 selectivity = selectivity,
                 Aoffset=df.sample$ageOffset,
                 start=list(Linf=1600,K=0.05,t0=-1.8,CV=0.1))
summary(fit.lsgm)
```

Plot the fitted growth curve, confidence intervals, and tolerance intervals for future observations
```{r}
d.lsgm <- predict(fit.lsgm,seq(0,maxAge,0.1),mean(season))
plot(length~I(age+ageOffset),data=df.sample,pch=16,col=grey(0.5,0.1),ylim=c(0,1.2*Linf))
lines(length~age,data=d.lsgm,col="dodgerblue4")
matlines(d.lsgm$age,d.lsgm$length+d.lsgm$se%o%c(-2,2),lty=1,col="dodgerblue1")
matlines(d.lsgm$age,d.lsgm$length+d.lsgm$sd%o%c(-2,2),lty=1,col="dodgerblue2")
lines(length~age,data=d.nlme,col="firebrick")
```


Profile plot and confidence interval for `Linf`
```{r}
pr <- profile(fit.lsgm,"Linf",trace=FALSE)
plot(pr)
confint(pr)
```

Profile plot and confidence interval for `K`
```{r}
pr <- profile(fit.lsgm,"K",trace=FALSE)
plot(pr)
confint(pr)
```



### LGSM - with ageing error

Fit a von Bertalanffy model allowing for length selection, but ignoring ageing area
```{r}
library(lsgm)
fit.lsgm <- lsVB(df.sample$age,df.sample$length,ageErr,breaks,frac,
                 selectivity = selectivity,
                 Aoffset=df.sample$ageOffset,
                 start=list(Linf=1600,K=0.05,t0=-1.8,CV=0.1))
summary(fit.lsgm)
```

Plot the fitted growth curve, confidence intervals, and tolerance intervals for future observations
```{r}
d.lsgm <- predict(fit.lsgm,seq(0,maxAge,0.1),mean(season))
plot(length~I(age+ageOffset),data=df.sample,pch=16,col=grey(0.5,0.1),ylim=c(0,1.2*Linf))
lines(length~age,data=d.lsgm,col="dodgerblue4")
matlines(d.lsgm$age,d.lsgm$length+d.lsgm$se%o%c(-2,2),lty=1,col="dodgerblue1")
matlines(d.lsgm$age,d.lsgm$length+d.lsgm$sd%o%c(-2,2),lty=1,col="dodgerblue2")
lines(length~age,data=d.nlme,col="firebrick")
```


Profile plot and confidence interval for `Linf`
```{r}
pr <- profile(fit.lsgm,"Linf",trace=FALSE)
plot(pr)
confint(pr)
```

Profile plot and confidence interval for `K`
```{r}
pr <- profile(fit.lsgm,"K",trace=FALSE)
plot(pr)
confint(pr)
```
